---
title: "Extracting climatic data - Population locations"
author: "Juliette Archambeau & Adélaïde Theraroz"
date: "`r format(Sys.time(), '%d %B, %Y')`"
number-sections: true
format: 
  html:
    toc: true
    toc-depth: 4
    code-fold: true
    page-layout: full
embed-resources: true
editor_options: 
  chunk_output_type: console
---

<style type="text/css">
body {
   font-size: 15px;
}
code.r{
  font-size: 11px;
}
pre {
  font-size: 11px
}

table {
  font-size: 10px
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache=F)
options(width = 300)
library(knitr)      # CRAN v1.26
library(kableExtra) # CRAN v1.1.0
library(tidyverse)  # CRAN v1.3.0
library(janitor)
library(purrr)
library(raster)
library(magrittr)
library(here)
library(readxl)

# my own functions
source(here("scripts/functions/kable_mydf.R")) # building tables
source(here("scripts/functions/extract_climatedt_metadata.R")) # extracting meta data of the climatic variables in ClimateDT
source(here("scripts/functions/extract_clim_from_rasters.R")) # extracting climatic values from rasters
```

# Set of climatic variables

The set of climatic variables has been selected in Archambeau et al. (2023): see the [associated Github repository](https://github.com/JulietteArchambeau/GOPredEvalPinpin) and [details on the climatic variable selection](https://juliettearchambeau.github.io/GOPredEvalPinpin/0_FormattingPopulationCoordinatesElevationClimateData.html). Here is the set of selected climatic variables:

```{r ListSelectedClimVariables}
clim_var <- c("bio1","bio12","bio15","bio3","bio4","SHM")

extract_climatedt_metadata(var_clim = clim_var) %>% 
  dplyr::select(label,description,unit) %>% 
  set_colnames(str_to_title(colnames(.))) %>% 
  kable_mydf(font_size = 12)
```

# Population coordinates

We load the dataset with the population coordinates and we create a column `pop` which corresponds to the first three letters of the population codes (column `code`).

**Warning:** 

  - In previous studies, the VAL-VMQ population in Spain was referred as VAL. However, in the present study, this population is named VMQ and the VAL population corresponds to the VAL-CORSICA population.
  
  - The MAU and MAUbis populations are very close, but with an altitude difference of almost 300m. We assign them different codes: MAU for the MAU population and MAB for the MAUbis population.

```{r LoadPopulationCoordinates}
pop_coord <-  read_csv(here("data/population_information_pinaster_feb2023.csv"),show_col_types = FALSE) %>% 
  mutate(pop = case_when(code=="VAL-VMQ" ~"VMQ",
                         code== "MAUbis/FR-PP-12" ~ "MAB",
                         TRUE ~ str_sub(code,1,3)))

pop_coord %>% kable_mydf()
```

This dataset contains `r nrow(pop_coord)` populations.

We are going to keep only the populations that will be used in the following analyses (i.e. GEAs and genomic offset estimation). For that, we load the population-level allele frequencies that will be used in the GEA analyses and we extract the population names:

```{r PopToKeep}
pop_tokeep <- readRDS(here("data/genomic_data/imputed_allele_frequencies.rds")) %>% pull(pop)
```

The genomic data contain `r length(pop_tokeep)` populations.

```{r FilterPops}
pop_coord <- pop_coord %>% 
  filter(pop %in% pop_tokeep) %>% 
  arrange(pop)

pop_coord %>% kable_mydf()
```

We create spatial points for the population coordinates:

```{r PopCoordToSpatialPoints}
pop_sp <- pop_coord %>% 
  dplyr::select(longitude,latitude) %>% 
  SpatialPoints(proj4string=CRS("+proj=longlat +datum=WGS84 +no_defs"))
```

# Extracting past climates

We extract the climatic variables for the reference period 1901-1950.

```{r ExtractRefenceClimate}
path_to_rasters <- "C:/Users/jularc/Documents/GOPredEvalPinpin/GOPredEvalPinpin/data/ClimaticData/ClimateDTRasters/"
  
  
clim_ref <- lapply(clim_var, function(x) {
  
  x %>% 
    extract_clim_from_rasters(x=x,
                              period="1901-1950",
                              pop_coord=pop_sp,
                              path=path_to_rasters)}) %>% 
  setNames(clim_var) %>% 
  as_tibble() %>% 
  mutate(pop=pop_coord$pop) %>% 
  dplyr::select(pop,everything())

clim_ref %>% kable_mydf()

clim_ref %>% saveRDS(here("data/climatic_data/reference_climate_population_locations.rds"))
```

# Extracting future climates 

We extract the future climate for the different global climate models (GCMs), for the Shared Socioeconomic Pathway (SSP) 3-7.0 and the reference period 2041-2070. 

Here are the GCM names:

```{r GCMnames}
gcm_names <- list.files(here(path_to_rasters)) %>%  str_sub(5,-35) %>% setdiff("")
gcm_names 
```


```{r ExtractingFutureClimate}
# List of dataframes (one for each GCM) with the climatic variables in columns
clim_fut <- lapply(gcm_names, function(gcm){
  
lapply(clim_var, function(x) extract_clim_from_rasters(x=x, 
                                                       gcm=gcm, 
                                                       period="2041-2070", 
                                                       ssp="ssp370",
                                                       pop_coord=pop_sp,
                                                       path=path_to_rasters)) %>% 
  setNames(clim_var) %>% 
  as_tibble() %>% 
  mutate(pop=pop_coord$pop,
         gcm=gcm) %>% 
  dplyr::select(pop,gcm,everything())

}) %>% 
  setNames(gcm_names)

clim_fut %>% saveRDS(here("data/climatic_data/future_climate_population_locations.rds"))
```

